`cesR`: A package to make using the Canadian Election Study a little easier

### 2020-08-27 paper.Rmd
updated formatting

---
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: false
    fig_caption: true
    number_sections: true
    latex_engine: pdflatex
title: "`cesR`: A package to make using the Canadian Election Study a little easier"
thanks: "We thank **THANKS**.  Our code and datasets are available at: **LINK**. Comments on the `r format(Sys.time(), '%d %B %Y')` version of this paper are welcome at: rohan.alexander@utoronto.ca. For the latest version please click [here](LINK)."
author:
- Paul Hodgetts, University of Toronto
- Rohan Alexander, University of Toronto
abstract: "TBD"
keywords: "TBD"
toc: false
date: "`r format(Sys.time(), '%d %B %Y')`"
bibliography: references.bib
biblio-style: "apalike"
endnote: no
---

@@

The CES survey datasets are publicly available from various repositories across the internet in various data types ([@ces2019]; [@ubc]; [@odesi]). As the ability to find and access data is integral to quality research in any domain, the `cesR` package brings together these various sources, providing R users with a means of easily accessing each CES survey dataset while also circumventing the need to download and store survey data files. The structure of `cesR` also removes the need to call a data file from a directory, thereby making it easier to work with CES datasets between different workstations. Additionally, `cesR` provides a function, `get_decon()`, that creates a subset of the 2019 CES online survey that has been prepared in an opinionated way. This subset of the 2019 CES online survey provides a tool for educators that can be used to aid in the teaching of statistical analysis of large survey datasets.

The `cesR` package is built using the R programming language and is designed to work within the R integrated development environment (IDE) RStudio ([@citeR]; [@citeRStudio]). Providing access to data is common in R packages and the `cesR` package follows and was inspired by the work being done in the R community through such packages as the `opendatatoronto` package [@citeopendatatoronto], the `Lahman` package [@citeLahman], `fueleconomy` [@citefuel], and `nasaweather` [@citenasa]. Furthermore, the `cesR` package is complementary to packages such as `dataverse` [@citedataverse] and `cancensus` [@citecancensus] that provide R users access to Canadian survey and census data. Packages such as these are important to R users as they improve the functionality of working within R by minimizing the number of steps required to load data and increasing the availability of data to R users. The `cesR` package contributes to this area of R package development and the utility of R by providing access to CES survey data while also providing a tool for educators.

@@

# convert dataframe values to factor type
# check column heads for dataframe
ces2019_web <- to_factor(ces2019_web)
head(ces2019_web)
```
 cps19_StartDate | cps19_EndDate | cps19_ResponseId | cps19_consent | cps19_citizensh~ | cps19_yob | cps19_yob_2001_~ | cps19_gender | cps19_province | cps19_eduction |
-----------------|---------------|------------------|---------------|------------------|-----------|------------------|--------------|----------------|----------------|
<dttm> | <dttm> | <chr> | <fct> | <fct> | <fct> | <fct> | <fct> | <fct> | <fct> |
2019-09-13 08:09:44 | 2019-09-13 08:36:19 | R_1OpYXEFGzHRUp~ | I consent to~ | Canadian citizen | 1989 | NA | A woman | Quebec | Master's degree |
2019-09-13 08:39:09 | 2019-09-13 08:57:06 | R_2qdrL3J618rxY~ | I consent to~ | Canadian citizen | 1998 | NA | A woman | Quebec | Master's degree |
2019-09-13 10:01:19 | 2019-09-13 10:27:29 | R_USWDAPcQEQiMm~ | I consent to~ | Canadian citizen | 2000 | NA | A woman | Ontario| Some university |
2019-09-13 10:05:37 | 2019-09-13 10:50:53 | R_3IQaeDXy0tBzE~ | I consent to~ | Canadian citizen | 1998 | NA | A man   | Ontario| Some university |
2019-09-13 10:05:52 | 2019-09-13 10:32:53 | R_27WeMQ1asip2c~ | I consent to~ | Canadian citizen | 2000 | NA | A woman | Ontario| Completed seco~ |
2019-09-13 10:10:20 | 2019-09-13 10:29:45 | R_3LiGZcCWJEcWV~ | I consent to~ | Canadian citizen | 1999 | NA | A woman | Ontario| Some university |

@@

# get preview for 2019 CES online survey for 10 rows
get_preview("ces2019_web", 10)
```
cps19_StartDate | cps19_EndDate | cps19_ResponseId | cps19_consent | cps19_citizensh~ | cps19_yob | cps19_yob_2001_~ | cps19_gender | cps19_province | cps19_eduction | cps19_demsat | cps19_imp_iss   |
----------------|---------------|------------------|---------------|------------------|-----------|------------------|--------------|----------------|----------------|--------------|-----|
<dttm>          |    <dttm>     |         <chr>    |         <fct> |         <fct>    |   <fct>   |  <fct>           |  <fct>       | <fct>          | <fct>          |  <fct>       |
<chr>           |
2019-09-13 08:09:44 | 2019-09-13 08:36:19 | R_1OpYXEFGzHRUp~ | I consent to~ | Canadian citizen | 1989 | NA | A woman | Quebec | Master's degree | Fairly sati~ | environnement |
2019-09-13 08:39:09 | 2019-09-13 08:57:06 | R_2qdrL3J618rxY~ | I consent to~ | Canadian citizen | 1998 | NA | A woman | Quebec | Master's degree | Don't know/~ | NA |
2019-09-13 10:01:19 | 2019-09-13 10:27:29 | R_USWDAPcQEQiMm~ | I consent to~ | Canadian citizen | 2000 | NA | A woman | Ontario| Some university | Fairly sati~ | NA |
2019-09-13 10:05:37 | 2019-09-13 10:50:53 | R_3IQaeDXy0tBzE~ | I consent to~ | Canadian citizen | 1998 | NA | A man   | Ontario| Some university | Fairly sati~ | NA |
2019-09-13 10:05:52 | 2019-09-13 10:32:53 | R_27WeMQ1asip2c~ | I consent to~ | Canadian citizen | 2000 | NA | A woman | Ontario| Completed seco~ | Fairly sati~ | NA |
2019-09-13 10:10:20 | 2019-09-13 10:29:45 | R_3LiGZcCWJEcWV~ | I consent to~ | Canadian citizen | 1999 | NA | A woman | Ontario| Some university | Fairly sati~ | NA |
2019-09-13 10:14:47 | 2019-09-13 10:32:32 | R_1Iu8R1UlYzVMy~ | I consent to~ | Permanent resid~ | 1998 | NA | A woman | Ontario| Completed seco~ | Fairly sati~ | NA |           
2019-09-13 10:15:39 | 2019-09-13 10:30:59 | R_2EcS26hqrcVYl~ | I consent to~ | Canadian citizen | 1998 | NA | A woman | Ontario| Some university | Fairly sati~ | NA |           
2019-09-13 10:15:48 | 2019-09-13 10:37:45 | R_3yrt44wqQ1d4V~ | I consent to~ | Canadian citizen | 1998 | NA | A woman | Ontario| Some university | Fairly sati~ | NA |           
2019-09-13 10:16:08 | 2019-09-13 10:40:14 | R_10OBmXJyvn8fe~ | I consent to~ | Canadian citizen | 2000 | NA | A man   | Ontario| Completed seco~ | Fairly sati~ | NA |

@@

# get preview for 2019 CES online survey
get_preview("ces2019_web")
```
cps19_StartDate | cps19_EndDate | cps19_ResponseId | cps19_consent | cps19_citizensh~ | cps19_yob | cps19_yob_2001_~ | cps19_gender | cps19_province | cps19_eduction | cps19_demsat | cps19_imp_iss   |
----------------|---------------|------------------|---------------|------------------|-----------|------------------|--------------|----------------|----------------|--------------|-----|
<dttm>          |    <dttm>     |         <chr>    |         <fct> |         <fct>    |   <fct>   |  <fct>           |  <fct>       | <fct>          | <fct>          |  <fct>       |
<chr>           |
2019-09-13 08:09:44 | 2019-09-13 08:36:19 | R_1OpYXEFGzHRUp~ | I consent to~ | Canadian citizen | 1989 | NA | A woman | Quebec | Master's degree | Fairly sati~ | environnement |
2019-09-13 08:39:09 | 2019-09-13 08:57:06 | R_2qdrL3J618rxY~ | I consent to~ | Canadian citizen | 1998 | NA | A woman | Quebec | Master's degree | Don't know/~ | NA |
2019-09-13 10:01:19 | 2019-09-13 10:27:29 | R_USWDAPcQEQiMm~ | I consent to~ | Canadian citizen | 2000 | NA | A woman | Ontario| Some university | Fairly sati~ | NA |
2019-09-13 10:05:37 | 2019-09-13 10:50:53 | R_3IQaeDXy0tBzE~ | I consent to~ | Canadian citizen | 1998 | NA | A man   | Ontario| Some university | Fairly sati~ | NA |
2019-09-13 10:05:52 | 2019-09-13 10:32:53 | R_27WeMQ1asip2c~ | I consent to~ | Canadian citizen | 2000 | NA | A woman | Ontario| Completed seco~ | Fairly sati~ | NA |
2019-09-13 10:10:20 | 2019-09-13 10:29:45 | R_3LiGZcCWJEcWV~ | I consent to~ | Canadian citizen | 1999 | NA | A woman | Ontario| Some university | Fairly sati~ | NA |

@@

 ces_code | citizenship | yob | gender | province_territory | education | lr | lr_bef | lr_aft | regligion | sexuality_select | sexuality_text | language_eng | language_fr | langauge_abgi | employment | income | income_cat| marital | econ_retro | econ_fed | econ_self |
----------|-------------|-----|--------|--------------------|-----------|----|--------|--------|-----------|------------------|----------------|--------------|-------------|---------------|------------|--------|-----------|----------|------------|----------|-----------|
ces2019_web|Canadian citizen|1989|A woman|Quebec|Master's degree|2|NA|2|Don't know/ Prefer not to answer|Prefer not to say|NA|NA|NA|NA|Don't know/ Prefer not to answer|NA|Don't know/ Prefer not to answer|Don't know/ Prefer not to answer|Stayed about the same|Don't know/ Prefer not to answer|Don't know/ Prefer not to answer|
ces2019_web|Canadian citizen|1998|A woman|Quebec|Master's degree|2|NA|2|Catholic/ Roman Catholic/ RC|Prefer not to say|NA|English|French|NA|Student and working for pay|NA|Don't know/ Prefer not to answer|Living with a partner|Stayed about the same|Not made much difference|Not made much difference|
ces2019_web|Canadian citizen|1998|A man|Ontario|Some university|7|7|NA|Jewish/ Judaism/ Jewish Orthodox|Heterosexual|NA|English|NA|NA|Student and working for pay|NA|$110,001 to $150,000|Never Married|Got worse|Worse|Not made much difference

@@

### Preview dataset
```
# get preview of first 8 rows of 2019 CES phone survey dataset
get_preview("ces2019_phone", 8)
```
 sample_id | survey_end_CES | survey_end_mont~ | survey_end_day_~ | num_attempts_CES | interviewer_id_~ | interviewer_gen~ | language_CES | phonetype_CES |
-----------|----------------|------------------|------------------|------------------|------------------|------------------|--------------|---------------|
<dbl>      | <chr>          | <dbl>            | <dbl>            |  <dbl>           |  <dbl>           |     <chr>        |    <fct>     | <fct>         |
18         | 2019-09-23 15~ |               9  |              23  |              5   |         161182   | Female           | (2) French   |  (2) Wireless |
32         | 2019-09-12 18~ |               9  |              12  |              1   |         151152   | Male             | (1) English  |  (2) Wireless |
39         | 2019-09-10 18~ |               9  |              10  |              1   |         161182   | Female           | (2) French   |  (2) Wireless |
59         | 2019-10-10 15~ |              10  |              10  |              6   |         147601   | Female           | (2) French   |  (2) Wireless |
61         | 2019-09-12 16~ |               9  |              12  |              1   |         151152   | Male             | (2) French   |  (2) Wireless |
69         | 2019-09-17 17~ |               9  |              17  |              1   |           2503   | Female           | (2) French   |  (2) Wireless |       
157        | 2019-09-12 16~ |               9  |              12  |              1   |           2529   | Male             |  (2) French  | (2) Wireless   | 
158        | 2019-09-14 10~ |                9 |              14  |               4  |         142515   | Male             |  (2) French  | (2) Wireless   |

@@

# alternatively the factored table can be assigned to its own data object to leave the original untouched.
# ces2019_phone_factor <- labelled::to_factor(ces2019_phone)
# head(ces2019_phone_factor)
```
 sample_id | survey_end_CES | survey_end_mont~ | survey_end_day_~ | num_attempts_CES | interviewer_id_~ | interviewer_gen~ | language_CES | phonetype_CES |
-----------|----------------|------------------|------------------|------------------|------------------|------------------|--------------|---------------|
<dbl>      | <chr>          | <dbl>            | <dbl>            |  <dbl>           |  <dbl>           |     <chr>        |    <fct>     | <fct>         |
18         | 2019-09-23 15~ |               9  |              23  |              5   |         161182   | Female           | (2) French   |  (2) Wireless |
32         | 2019-09-12 18~ |               9  |              12  |              1   |         151152   | Male             | (1) English  |  (2) Wireless |
39         | 2019-09-10 18~ |               9  |              10  |              1   |         161182   | Female           | (2) French   |  (2) Wireless |
59         | 2019-10-10 15~ |              10  |              10  |              6   |         147601   | Female           | (2) French   |  (2) Wireless |
61         | 2019-09-12 16~ |               9  |              12  |              1   |         151152   | Male             | (2) French   |  (2) Wireless |
69         | 2019-09-17 17~ |               9  |              17  |              1   |           2503   | Female           | (2) French   |  (2) Wireless |

@@

head(ces2019_phone_subset)
```
interviewer_gender~ |  language_CES | phonetype_CES | weight_CES | q1      | q2      | q3       | q4        | q6            | q7     |
--------------------|---------------|---------------|------------|---------|---------|----------|-----------|---------------|--------|
 <chr>              |     <dbl+lbl> |   <dbl+lbl>   |   <dbl>    |<dbl+lbl>| <dbl+l> | <dbl+lb> | <dbl+lbl> |  <dbl+lbl>    | <chr>  |
Female              |2 [(2) French] |2 [(2) Wireles~|      0.902 |1 [(1) Y~|    1963 |1 [(1) ~  |5 [(5) Q~  |3 [(3) Not ve~ |economie|
Male                |1 [(1) Englis~ |2 [(2) Wireles~|      0.902 |1 [(1) Y~|    1973 |1 [(1) ~  |5 [(5) Q~  |2 [(2) Fairly~ |Finances|
Female              |2 [(2) French] |2 [(2) Wireles~|      0.902 |1 [(1) Y~|    1994 |1 [(1) ~  |5 [(5) Q~  |1 [(1) Very s~ |agriculture|
Female              |2 [(2) French] |2 [(2) Wireles~|      1.23  |1 [(1) Y~|    2000 |1 [(1) ~  |5 [(5) Q~  |2 [(2) Fairly~ |l'environnement|
Male                |2 [(2) French] |2 [(2) Wireles~|      0.902 |1 [(1) Y~|    1984 |1 [(1) ~  |5 [(5) Q~  |4 [(4) Not sa~ |Utilisation de la Banque du ~|
Female              |2 [(2) French] |2 [(2) Wireles~|      0.902 |1 [(1) Y~|    1939 |1 [(1) ~  |5 [(5) Q~  |3 [(3) Not ve~ |Notre identité Québécoise|

>>>>

---
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: false
    fig_caption: true
    number_sections: true
    latex_engine: pdflatex
title: "`cesR`: A package to make using the Canadian Election Study a little easier"
thanks: "We thank **THANKS**.  Our code and datasets are available at: **LINK**. Comments on the `r format(Sys.time(), '%d %B %Y')` version of this paper are welcome at: rohan.alexander@utoronto.ca. For the latest version please click [here](LINK)."
author:
- Paul A. Hodgetts, University of Toronto
- Rohan Alexander, University of Toronto
abstract: "TBD"
keywords: "TBD"
toc: false
date: "`r format(Sys.time(), '%d %B %Y')`"
bibliography: references.bib
biblio-style: "apalike"
endnote: no
---

@@

The CES survey datasets are publicly available from various repositories across the internet in various data types [@ces2019; @ubc; @odesi]. As the ability to find and access data is integral to quality research in any domain, the `cesR` package brings together these various sources, providing R users with a means of easily accessing each CES survey dataset while also circumventing the need to download and store survey data files. The structure of `cesR` also removes the need to call a data file from a directory, thereby making it easier to work with CES datasets between different workstations. Additionally, `cesR` provides a function, `get_decon()`, that creates a subset of the 2019 CES online survey that has been prepared in an opinionated way. This subset of the 2019 CES online survey provides a tool for educators that can be used to aid in the teaching of statistical analysis of large survey datasets.

The `cesR` package is built using the R programming language and is designed to work within the R integrated development environment (IDE) RStudio [@citeR; @citeRStudio]. Providing access to data is common in R packages and the `cesR` package follows and was inspired by the work being done in the R community through such packages as the `opendatatoronto` package [@citeopendatatoronto], the `Lahman` package [@citeLahman], `fueleconomy` [@citefuel], and `nasaweather` [@citenasa]. Furthermore, the `cesR` package is complementary to packages such as `dataverse` [@citedataverse] and `cancensus` [@citecancensus] that provide R users access to Canadian survey and census data. Packages such as these are important to R users as they improve the functionality of working within R by minimizing the number of steps required to load data and increasing the availability of data to R users. The `cesR` package contributes to this area of R package development and the utility of R by providing access to CES survey data while also providing a tool for educators.

@@

# convert dataframe values to factor type
# check column heads for dataframe
ces2019_web <- to_factor(ces2019_web)
head(ces2019_web)

>head(ces2019_web)
# A tibble: 6 x 620
  cps19_StartDate     cps19_EndDate       cps19_ResponseId cps19_consent cps19_citizensh~ 
  <dttm>              <dttm>              <chr>            <fct>         <fct>               
1 2019-09-13 08:09:44 2019-09-13 08:36:19 R_1OpYXEFGzHRUp~ I consent to~ Canadian citizen      
2 2019-09-13 08:39:09 2019-09-13 08:57:06 R_2qdrL3J618rxY~ I consent to~ Canadian citizen      
3 2019-09-13 10:01:19 2019-09-13 10:27:29 R_USWDAPcQEQiMm~ I consent to~ Canadian citizen      
4 2019-09-13 10:05:37 2019-09-13 10:50:53 R_3IQaeDXy0tBzE~ I consent to~ Canadian citizen        
5 2019-09-13 10:05:52 2019-09-13 10:32:53 R_27WeMQ1asip2c~ I consent to~ Canadian citizen      
6 2019-09-13 10:10:20 2019-09-13 10:29:45 R_3LiGZcCWJEcWV~ I consent to~ Canadian citizen 
```

@@

# get preview for 2019 CES online survey for 10 rows
get_preview("ces2019_web", 10)
```

@@

# get preview for 2019 CES online survey
get_preview("ces2019_web")
```
cps19_StartDate | cps19_EndDate | cps19_ResponseId | cps19_consent | cps19_citizensh~ |
-----|-----|-------|----|-----|
2019-09-13 08:09:44 | 2019-09-13 08:36:19 | R_1OpYXEFGzHRUp~ | I consent to~ | Canadian citizen |
2019-09-13 08:39:09 | 2019-09-13 08:57:06 | R_2qdrL3J618rxY~ | I consent to~ | Canadian citizen |
2019-09-13 10:01:19 | 2019-09-13 10:27:29 | R_USWDAPcQEQiMm~ | I consent to~ | Canadian citizen |
2019-09-13 10:05:37 | 2019-09-13 10:50:53 | R_3IQaeDXy0tBzE~ | I consent to~ | Canadian citizen |
2019-09-13 10:05:52 | 2019-09-13 10:32:53 | R_27WeMQ1asip2c~ | I consent to~ | Canadian citizen |
2019-09-13 10:10:20 | 2019-09-13 10:29:45 | R_3LiGZcCWJEcWV~ | I consent to~ | Canadian citizen |

@@

LINK: https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/DUS88V
```
```
# A tibble: 6 x 22
  ces_code citizenship yob   gender province_territ~ education lr    lr_bef lr_aft religion 
  <chr>    <fct>       <fct> <fct>  <fct>            <fct>     <chr> <chr>  <chr>  <fct>           
1 ces2019~ Canadian c~ 1989  A wom~ Quebec           Master's~ 2     NA     2      Don't k~           
2 ces2019~ Canadian c~ 1998  A wom~ Quebec           Master's~ 2     NA     2      Catholi~      
3 ces2019~ Canadian c~ 2000  A wom~ Ontario          Some uni~ 4     4      NA     Catholi~      
4 ces2019~ Canadian c~ 1998  A man  Ontario          Some uni~ 7     7      NA     Jewish/~      
5 ces2019~ Canadian c~ 2000  A wom~ Ontario          Complete~ 4     4      NA     Catholi~      
6 ces2019~ Canadian c~ 1999  A wom~ Ontario          Some uni~ 4     4      NA     Catholi~ 
```

@@

### Preview dataset
```
# get preview of first 8 rows of 2019 CES phone survey dataset
get_preview("ces2019_phone", 8)
```
 sample_id | survey_end_CES | survey_end_mont~ | survey_end_day_~ | num_attempts_CES |
---|------|-------|------|------|---|----|
<dbl>      | <chr>          | <dbl>            | <dbl>            |  <dbl>           |
18         | 2019-09-23 15~ |               9  |              23  |              5   |
32         | 2019-09-12 18~ |               9  |              12  |              1   |
39         | 2019-09-10 18~ |               9  |              10  |              1   |
59         | 2019-10-10 15~ |              10  |              10  |              6   |
61         | 2019-09-12 16~ |               9  |              12  |              1   |
69         | 2019-09-17 17~ |               9  |              17  |              1   |
157        | 2019-09-12 16~ |               9  |              12  |              1   |
158        | 2019-09-14 10~ |               9  |              14  |              4   |

@@

# alternatively the factored table can be assigned to its own data object 
# to leave the original untouched.
# ces2019_phone_factor <- labelled::to_factor(ces2019_phone)
# head(ces2019_phone_factor)

>head(ces2019_phone)
# A tibble: 6 x 278
  sample_id survey_end_CES survey_end_mont~ survey_end_day_~ num_attempts_CES 
      <dbl> <chr>                     <dbl>            <dbl>            <dbl> 
1        18 2019-09-23 15~                9               23                5 
2        32 2019-09-12 18~                9               12                1 
3        39 2019-09-10 18~                9               10                1 
4        59 2019-10-10 15~               10               10                6 
5        61 2019-09-12 16~                9               12                1 
6        69 2019-09-17 17~                9               17                1 
```

@@

head(ces2019_phone_subset)

>head(ces2019_phone_subset)
# A tibble: 6 x 10
  interviewer_gender~ language_CES phonetype_CES weight_CES q1     q2    q3      q4                                    
  <chr>               <fct>        <fct>              <dbl> <fct>  <fct> <fct>   <fct>    
1 Female              (2) French   (2) Wireless       0.902 (1) Y~ 1963  (1) Ma~ (5) Que~
2 Male                (1) English  (2) Wireless       0.902 (1) Y~ 1973  (1) Ma~ (5) Que~
3 Female              (2) French   (2) Wireless       0.902 (1) Y~ 1994  (1) Ma~ (5) Que~
4 Female              (2) French   (2) Wireless       1.23  (1) Y~ 2000  (1) Ma~ (5) Que~
5 Male                (2) French   (2) Wireless       0.902 (1) Y~ 1984  (1) Ma~ (5) Que~
6 Female              (2) French   (2) Wireless       0.902 (1) Y~ 1939  (1) Ma~ (5) Que~
```
